BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2016-12-26 Paper/BOAST.tex
introduction

\section{Introduction}

Porting and tuning HPC applications to new platforms is tedious and costly in
term of human resources.  Many developer teams are more concentrated on solving
the physical problems faster and better with new algorithms and simulating new
physics. This is why optimization, while important, is a task which is
sometimes neglected or hidden for many developers.  In the same way, Domain
Specific Language (DSLs) have exactly this goal, the complete separation of the
high-level code dedicated to simulate new physics and the low-level one which
is more dependent on the HPC architecture. Nevertheless, decoupling the high
and low levels don't avoid the need of optimization which becomes more and more
unavoidable with new architectures based on many cores and accelerators.

Unfortunately, portability efforts of the low level i.e. computing kernels are
often lost when migrating to a new architecture. Worse, code may lose
maintainability because several versions of some functionalities coexist,
usually with a lot of duplication.  Thus productivity of porting and tuning
efforts is low as a huge fraction of those developments are never used after
the platform they were intended for is decommissioned.

The primary goal of the BOAST metaprogramming framework is to simplify the
optimization of computing kernels for HPC applications. The main idea is to use
a high level language to define an API of computing kernels i.e.  subroutines
and their related arguments, but also describe the way to optimize it
(unrolling, vectorization, parametrization, templating, tiling).  BOAST will
test automatically the many possibilities of optimizations and choose the best
one after validation from a reference version.  Then BOAST can generate pieces
of Fortran, C, OpenCL, or CUDA codes, test it for a specific target
architecture.

With the definition of an API, BOAST can hide the implementation details at
high level but also is a way to factorize different kernels depending only on
parameters such as data types, length of filters or boundary conditions.  The
maintainability and genericity of HPC codes is often limited. One of the reason
is that producing generic code in Fortran 90/95 is difficult as the language
does not really fit for it. Sometimes, adding genericity or functionalities can
be quite costly and degrades performance as optimization opportunities that
come from over-specification are lost. BOAST can provide a solution to handle
genericity and add new functionalities at low levels.

@@

The paper is organized as follows. First we further motivate the need for
metaprogramming frameworks.  Secondly we illustrate the optimization problems
related to scientific computing with the Laplace equation. Thirdly we present
our solution, called BOAST, its architecture and its functionalities. Fourthly
we apply BOAST to three use cases: the Laplace equation motivating example;
short multi-dimensional convolutions from the BigDFT quantum chemistry
application; and the seismic wave propagation simulator SPECFEM3D where 80 GPU
computing kernels have been ported improving, considerably the maintainability
and providing a single code base for OpenCL and CUDA kernel versions. Finally
we compare BOAST with other related works.

\section{Background and Motivation}

@@

  \subsection{Evolution of HPC Architectures}

@@

\section{Basic Example: OpenCL Laplace}
\label{sec:laplace}

@@

\section{BOAST: Using Code Generation in Application Autotuning}

>>>>

%%% INTRODUCTION %%%
\section{Introduction}

Porting and tuning HPC applications to new platforms is tedious and costly in term of human resources.  Many developer teams are more concentrated on
solving the physical problems faster and better with new algorithms and simulating new physics. This is why optimization, while important, is a task
which is sometimes neglected or hidden for many developers.  In the same way, Domain Specific Language (DSLs) have exactly this goal, the complete
separation of the high-level code dedicated to simulate new physics and the low-level one which is more dependent on the HPC
architecture. Nevertheless, decoupling the high and low levels does not avoid the need of optimizations which become more and more unavoidable with
new architectures based on many cores and accelerators.

Unfortunately, portability efforts of the low level codes (i.e. computing kernels) are often lost when migrating to a new architecture. Worse, code
may lose maintainability because several versions of some functionalities coexist, usually with a lot of duplication.  Thus productivity of porting
and tuning efforts is low as a huge fraction of those developments are never used after the platform they were intended for is decommissioned.

The primary goal of the BOAST metaprogramming framework is to simplify the optimization of computing kernels for HPC applications. The main idea is to
use a high level language to define an API of computing kernels i.e. subroutines and their related arguments, but also describe the way to optimize
it (unrolling, vectorization, parametrization, templating, tiling). BOAST will then test automatically the many possibilities of optimizations and
choose the best one after validation from a reference version. Then BOAST can generate pieces of Fortran, C, OpenCL, or CUDA codes, test them for a
specific target architecture.

With the definition of an API, BOAST can hide the implementation details at high level but also is a way to factorize different kernels depending only
on parameters such as data types, length of filters, or boundary conditions. The maintainability and genericity of HPC codes is often limited. One of
the reason is that producing generic code in Fortran 90/95 is difficult as the language does not really fit for it. Sometimes, adding genericity or
functionalities can be quite costly and may degrade performance as optimization opportunities that come from over-specification are lost. BOAST can
provide a solution to handle genericity and add new functionalities at low levels.

@@

The paper is organized as follows. First in Section~\ref{sec:backg} we further motivate the need for metaprogramming frameworks.  Secondly in
Section~\ref{sec:laplace} we illustrate the optimization problems related to scientific computing with the Laplace equation. Thirdly we present our
solution, called BOAST, its architecture and its functionalities in Section~\ref{sec:boast}. Fourthly we apply BOAST to three use cases: the Laplace
equation motivating example, short multi-dimensional convolutions from the BigDFT quantum chemistry application, and the seismic wave propagation
simulator SPECFEM3D where 80 GPU computing kernels have been ported improving considerably the maintainability and providing a single code base for
OpenCL and CUDA kernel versions. Finally, before a conclusion and future works, we compare BOAST with other related works.

%%% BACKGROUND AND MOTIVATION %%%
\section{Background and Motivation\label{sec:backg}}

@@

%%%
\subsection{Evolution of HPC Architectures}

@@

%%%
\section{Basic Example: OpenCL Laplace\label{sec:laplace}}

@@

\section{BOAST: Using Code Generation in Application Autotuning\label{sec:boast}}
